---
title: "PUTTING IT ALL TOGETHER"
subtitle: "| NU Pathology LearnR! Fall 2020 \n| Session 6\n"
output:
  html_document:
    df_print: paged
---

```{r setup{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(janitor)
library(kableExtra)
library(ggpubr)
```


This report represents analyses conducted in Session 6 of the Fall 2020 NU Pathology LearnR! course. The goal is to replicate parts of Table 1 and Figure 1 from the following publication, using the associated, publicly available data set.  

*Source: LONG-TERM EFFECTS OF BUDESONIDE OR NEDOCROMIL IN CHILDREN WITH ASTHMA by THE CHILDHOOD ASTHMA MANAGEMENT PROGRAM RESEARCH GROUP published in N Engl J Med 2000;343:1054-63.*  

"The Childhood Asthma Management Program (CAMP) was a clinical trial carried out in children with asthma. The trial was designed to determine the long-term effects of 3 treatments (budesonide, nedocromil, or placebo) on pulmonary function as measured by normalized FEV1 over a 5-6.5 year period."  

"Conclusions: In children with mild-to-moderate asthma, neither budesonide nor nedocromil is better than placebo in terms of lung function, but inhaled budesonide improves airway responsiveness and provides better control of asthma than placebo or nedocromil. The side effects of budesonide are limited to a small, transient reduction in growth velocity."  

```{r load data, include = FALSE, message = FALSE}

camp_data <- read_csv("data/camp_teach.csv") %>%
              mutate_if(is.character, as.factor) %>%
              mutate_at(vars(anypet:fdays), as.factor) %>%
              mutate(TG = recode(TG, 
                                 "A" = "Budesonide",
                                 "B" = "Nedocromil",
                                 "C" = "Placebo")) %>%
              clean_names()
glimpse(camp_data)
```

```{r clean data, include = FALSE}

# Explore the dataset to identify problems
summary(camp_data)

# We can see that a number of variables have missing values (displayed as "NA")
# We can either drop these rows, replace the missing values with mean, or convert
# missing factor values to "explicit" missing values, i.e. they show up in figures
# and summaries as their own "Missing" category.

# Do the cleaning

# Replace missing values with mean (numeric) or explicit "Missing" coding (factor)
camp_data_no_miss <- camp_data %>%
                     mutate_if(is.numeric, funs(if_else(is.na(.), mean(., na.rm = T), .))) %>%
                     mutate_if(is.factor, funs(fct_explicit_na(.)))

summary(camp_data_no_miss)


# Assume weird outlier value '999' in hemog represents missings, replace with column mean
# Also see strange negative value in prefev -- possibly a key stroke error, as negative value does not make sense, replace with column mean, else leave as is
camp_data_clean <- camp_data_no_miss %>%
                   mutate(hemog = na_if(hemog, 999.0)) %>%
                   mutate(hemog = replace_na(hemog, 
                                             mean(hemog, na.rm = T))) %>%
                   mutate(prefev = ifelse(prefev < 0, mean(prefev), prefev))

summary(camp_data_clean)

```

```{r table instr, include = FALSE}
## Table 1: CHARACTERISTICS OF THE PATIENTS AT BASE LINE

# a) filter data for base line (visitc == 0)  
# b) do group_by treatment group, summarize for:  
#  -- age - yr (mean ± SD)  
#  -- Race or ethnic group — no. (%)   
#  -- Sex — no. (%)  
#  -- FEV1 before bronchodilator use — % of predicted ± SD  
#  -- FEV1 after bronchodilator use — % of predicted ± SD  
# c) fix labels to human readable vs column names and reorder, using recode_factor  
# d) have to include a new column for subgroup for formatting table as desired -- will use kableExtra package to group rows to create a single table
```


```{r table age, include = FALSE}
table1_age <- camp_data_clean %>%
            filter(visitc == 0) %>%
            group_by(tg) %>%
            summarize(age_mean = round(mean(age_rz),1), 
                      age_sd = round(sd(age_rz),1)) %>%
            unite(age_mean_sd, age_mean, age_sd, sep = "±") %>%
            spread(key = tg, value = age_mean_sd) %>%
            mutate(Subgroup = "", 
                   Characteristic = "Age — yr") 
```            
            
```{r table ethnic, include = FALSE}
table1_ethnic <- camp_data_clean %>%
                  filter(visitc == 0) %>%
                  mutate(ethnic = recode_factor(ethnic, 
                                "w" = "Non-Hispanic white",
                                "b" = "Non-Hispanic black",
                                "h" = "Hispanic",
                                "o" = "Other")) %>%
                  group_by(tg, ethnic) %>%
                  summarize(n = n()) %>%
                  mutate(pct = paste("(",round(100*n/sum(n),0), "%)", 
                                     sep = "")) %>%
                  unite(ethnic_n_pct, n, pct, sep = " ") %>%                
                  spread(key = tg, value = ethnic_n_pct) %>%
                  mutate(Subgroup = ethnic, 
                         Characteristic = "Race or ethnic group — no. (%)") %>%
                  select(-ethnic)
```

```{r table gender, include = FALSE}
table1_gender <- camp_data_clean %>%
                  filter(visitc == 0) %>%
                  mutate(gender = recode_factor(gender, 
                                "f" = "Female",
                                "m" = "Male")) %>%
                  group_by(tg, gender) %>%
                  summarize(n = n()) %>%
                  mutate(pct = paste("(",round(100*n/sum(n),0), "%)", 
                                     sep = "")) %>%
                  unite(gender_n_pct, n, pct, sep = " ") %>%                
                  spread(key = tg, value = gender_n_pct) %>%
                  mutate(Subgroup = gender,
                         Characteristic = "Gender — no. (%)") %>%
                  select(-gender)
```

```{r table FEV1 before, include = FALSE}
table1_prefev <- camp_data_clean %>%
            filter(visitc == 0) %>%
            group_by(tg) %>%
            summarize(prefev_mean = round(mean(prefev, na.rm = TRUE),1), 
                      prefev_sd = round(sd(prefev, na.rm = TRUE),1)) %>%
            unite(prefev_mean_sd, prefev_mean, prefev_sd, sep = "±") %>%
            spread(key = tg, value = prefev_mean_sd) %>%
            mutate(Subgroup = "",
                   Characteristic = "FEV1 before bronchodilator use — % of predicted")
```

```{r table FEV1 after, include = FALSE}
table1_posfev <- camp_data_clean %>%
            filter(visitc == 0) %>%
            group_by(tg) %>%
            summarize(posfev_mean = round(mean(posfev, na.rm = TRUE),1), 
                      posfev_sd = round(sd(posfev, na.rm = TRUE),1)) %>%
            unite(posfev_mean_sd, posfev_mean, posfev_sd, sep = "±") %>%
            spread(key = tg, value = posfev_mean_sd) %>%
            mutate(Subgroup = "",
                   Characteristic = "FEV1 after bronchodilator use — % of predicted")
```

```{r rbind table, include = FALSE}
full_table1 <- bind_rows(table1_age, table1_ethnic, table1_gender,
                        table1_prefev, table1_posfev)

```


```{r full table, echo = FALSE}
kable(select(full_table1, -Characteristic), 
      caption = "TABLE 1. CHARACTERISTICS OF THE PATIENTS AT BASE LINE.") %>%
  kable_styling() %>%
  pack_rows("Age — yr", 1, 1) %>%
  pack_rows("Race or ethnic group — no. (%)", 2, 5) %>%
  pack_rows("Gender — no. (%)", 6, 7) %>%
  pack_rows("FEV1 before bronchodilator use — % of predicted", 8, 8) %>%
  pack_rows("FEV1 after bronchodilator use — % of predicted", 9, 9)
```


```{r fig1 instructions, include = FALSE}
## Figure 1  
# Recreate the first 6 tiled line plots from Figure 1 of the paper, which shows means at annual intervals for each treatment group for each variable.
# 
# *Figure 1. Mean Values for Spirometric Measures before and after the Use of a Bronchodilator, Airway Responsiveness, Standing Height, and Standing-Height Velocity during Four Years of Follow-up in the Budesonide (Bud), Nedocromil (Ned), and Placebo (Plbo) Groups.*
# 
# *The numbers of observations used to calculate means at annual intervals are shown below each panel. When comparisons were made over the total follow-up time, the budesonide group differed significantly (P<0.001) from the placebo group in all measures, even though these differences may not be apparent in every panel, and there were no significant differences between the nedocro- mil group and the placebo group in any measure. FEV1 denotes forced expiratory volume in one second, FVC forced vital capacity, and FEV1 PC20 airway responsiveness measured by the concentration of methacholine that caused a 20 percent decrease in FEV1. For FEV1 PC20, values were obtained at 0, 8, 20, 32, and 44 months.*
```


```{r plot data prep, include = FALSE}
# filter for data at yearly visits for first 4 years (visitc %in% c(0, 12, 24, 36, 48))

camp_data_y04 <- camp_data_clean %>%
                  filter(visitc %in% c(0, 12, 24, 36, 48)) %>%
                  select(tg, visitc, posfevpp, posfev, posff, 
                         prefevpp, prefev, preff)

yr_means <- camp_data_y04 %>%
              group_by(tg, visitc) %>%
              summarize_at(vars(posfevpp:preff), 
                           mean, na.rm = TRUE)
```

```{r plots, include = FALSE}

#make the individual plots
#note the x-axis has to be numeric for line plot - must convert to character and then to numeric
# make it publication ready - fix labels and font sizes, theme, etc.
# we'll then assemble the individual plots into a single figure using a new package, ggpubr

#AFTER BD plots
p_posfevpp <- ggplot(yr_means, 
                     aes(x = as.numeric(as.character(visitc))/12, 
                         y = posfevpp)) +
              geom_line(aes(linetype = tg)) +
              labs(x = "Time (yr)", y = "% of Predicted Value", 
                   title = "FEV1 after Bronchodilator",
                   linetype = "Treatment group") +
              theme_bw() + 
              theme(legend.position = "bottom",
                    axis.title.x = element_text(size = 9),
                    axis.title.y = element_text(size = 9),
                    plot.title = element_text(size = 9))

p_posfev <- ggplot(yr_means, 
                     aes(x = as.numeric(as.character(visitc))/12, 
                         y = posfev)) +
              geom_line(aes(linetype = tg)) +
              labs(x = "Time (yr)", y = "Liters", 
                   title = "FEV1 after Bronchodilator",
                   linetype = "Treatment group") +
              theme_bw() + 
              theme(legend.position = "bottom",
                    axis.title.x = element_text(size = 9),
                    axis.title.y = element_text(size = 9),
                    plot.title = element_text(size = 9))

p_posff <- ggplot(yr_means, 
                     aes(x = as.numeric(as.character(visitc))/12, 
                         y = posff)) +
              geom_line(aes(linetype = tg)) +
              labs(x = "Time (yr)", y = "Ratio (percent)", 
                   title = "FEV1:FVC after Bronchodilator",
                   linetype = "Treatment group") +
              theme_bw() + 
              theme(legend.position = "bottom",
                    axis.title.x = element_text(size = 9),
                    axis.title.y = element_text(size = 9),
                    plot.title = element_text(size = 9))

#BEFORE BD plots
p_prefevpp <- ggplot(yr_means, 
                     aes(x = as.numeric(as.character(visitc))/12, 
                         y = prefevpp)) +
              geom_line(aes(linetype = tg)) +
              labs(x = "Time (yr)", y = "% of Predicted Value", 
                   title = "FEV1 before Bronchodilator",
                   linetype = "Treatment group") +
              theme_bw() + 
              theme(legend.position = "bottom",
                    axis.title.x = element_text(size = 9),
                    axis.title.y = element_text(size = 9),
                    plot.title = element_text(size = 9))

p_prefev <- ggplot(yr_means, 
                     aes(x = as.numeric(as.character(visitc))/12, 
                         y = prefev)) +
              geom_line(aes(linetype = tg)) +
              labs(x = "Time (yr)", y = "Liters", 
                   title = "FEV1 before Bronchodilator",
                   linetype = "Treatment group") +
              theme_bw() + 
              theme(legend.position = "bottom",
                    axis.title.x = element_text(size = 9),
                    axis.title.y = element_text(size = 9),
                    plot.title = element_text(size = 9)) 

p_preff <- ggplot(yr_means, 
                     aes(x = as.numeric(as.character(visitc))/12, 
                         y = preff)) +
              geom_line(aes(linetype = tg)) +
              labs(x = "Time (yr)", y = "Ratio (percent)", 
                   title = "FEV1:FVC before Bronchodilator",
                   linetype = "Treatment group") +
              theme_bw() + 
              theme(legend.position = "bottom",
                    axis.title.x = element_text(size = 9),
                    axis.title.y = element_text(size = 9),
                    plot.title = element_text(size = 9))
```

Figure 1. Mean Values for Spirometric Measures before and after the Use of a Bronchodilator during Four Years of Follow-up in the Budesonide (Bud), Nedocromil (Ned), and Placebo (Plbo) Groups.  
```{r grid plots, echo = FALSE}
ggarrange(p_posfevpp, p_posfev, p_posff, p_prefevpp, p_prefev, p_preff, 
          ncol = 3, nrow = 2,
          common.legend = TRUE, legend="bottom")

ggsave("images/Figure1.png", width = 7, height = 4, units = "in")
```

