---
title: "EXERCISES"
subtitle: "| NU Pathology LearnR! Fall 2020 \n| Session 1\n"
output:
  html_document:
    theme: cosmo
    toc: yes
    toc_float: yes
  html_notebook: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

```

---

## INTRO TO R & RSTUDIO

### PRACTICE: Writing R Code 
(a) Describe what is wrong with the code in the chunk below.   
**Solution: Variable names cannot start with numbers. A # is needed to comment out the sentence describing the code.**

(b) Fix the code to remove errors and use snake case for variables.  

```{r, eval = FALSE}
10.mean <- mean(rnorm(10))

This code calculates the mean of a series of random numbers.

```

**Solution:**
```{r}
mean_10 <- mean(rnorm(10))
#This code calculates the mean of a series of random numbers.
```

### PRACTICE: Writing R Code 
(a) Create a variable, `loc` with the following values: main, WST, NLX, and CD.   


(b) Add another value, GLB.    

**Solution:**
```{r }
loc <- c("main", "WST", "NLX", "CD")
loc <- c(loc, "GLB")

```
  
### PRACTICE: Loading and Using Packages 
(a) Identify what is wrong within the following code chunk.  
**Solution: You must first install the package and the argument for install.packages is a string.**

(b) Fix the code to properly load the package.  

```{r, eval = FALSE}

library(lubridate)
install.packages(lubridate)

```

**Solution:**
```{r, eval = FALSE}

install.packages("lubridate")
library(lubridate)

```
  
### PRACTICE: Using the help files 
Use a help file to examine the arguments for `read_csv()`.  

(a) What is the default setting for the skip argument?  
**Solution: The default setting is skip = 0** 


(b) What does the skip argument do?
**Solution: Number of lines to skip before reading data.** 



### EXTEND: Exploring the realm of available R packages
This website lists all packages available on CRAN: https://cran.r-project.org/web/packages/available_packages_by_name.html.   
  
(a) Using the search function in the Packages tab of RStudio to find packages loaded onto your computer, find and list at least one R package that can be used to read rectangular text files (.csv or .tsv) into R.  

Reads text files, is loaded on my computer:   

**Solution: readr**

(b) Now use an internet search to find an R package that can be used to read Excel files (.xls or .xlxs), but is not loaded onto your computer:  

Reads Excel files, is NOT loaded on my computer:  

**Solution: readxl or xlsx**
  
Select one of the packages and use its reference manual to find the documentation on the function to read a .xls file.    

(c) Write sample code below that would install the package, load it, and then use a function to read in a .xls file, titled ex1.xls (NOTE: Since there is no such Excel file, this code will not actually work, but is used for demonstrating correct syntax.)  

**Solution using readxl**  

```{r, eval = FALSE}

install.packages("readxl")
library(readxl)

ex1_data <- read_excel("ex1.xls")

#alternatively, this also works
ex1_data <- read_xls("ex1.xls")

# From the package documentation: read_excel() calls excel_format() to determine if path is xls or xlsx, based on the file extension and the file itself, in that order. Use read_xls() and read_xlsx() directly if you know better and want to prevent such guessing.

```

---

